BIOL 274 Homework 3

### 2024-02-03 bio_dynamics/bio274_hw3.Rmd
finished bio hw3

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(PNWColors)
```

@@

@@

@@

$$y(t)= 1250+C_2e^{-0.4t}$$

Check by differentiating:
$$y'(t)=-0.4\left(C_2e^{-0.4t}\right)$$
$$500-0.4y=500-0.4 \left(1250+C_2e^{-0.4t}\right)=$$
$$500-500-0.4\left(C_2e^{-0.4t}\right)=-0.4\left(C_2e^{-0.4t}\right)=y'(t)$$
So we have confirmed that out solution satisfies the original differential equation.

@@

For problems 3 and 4, find the solution to each of the following IVPs:
## Question 3
$$ \frac{dy}{dx}= \frac{-y}{x}+2, \quad y(1)=3 $$

@@

Check by differentiating:
$$p'(t)=(t+1)(4t)+(2t^2+3)=4t^2+4t+2t^2+3 $$
And since 
$$ 2t^2+3 =\frac{(t+1)(2t^2+3)}{t+1} = \frac{p(t)}{t+1}$$,

This means our solution indeed satisfies the IVP.

@@

```{r}
l_per_g <- 0.2641720524 # 1 L = 0.2641720524 gal
c <- 1/l_per_g
c
ppb5 <- 5*c

print(paste("1 part per billion = 1 µg per liter = 1 µg per", l_per_g,"gal = ", c, "µg/gal"))
print(paste("5 ppb =", ppb5, "µg/gal"))
```

<!-- $1 \text{ ppb} = \frac{1 \mu \text{g}} {\text{L}} = \frac{1 \mu\text{g}}{0.2641720524 \text{ gal}}= `r c` \mu \text{g/gal}$ -->

@@

$$\left(\frac{18.927 \text{ µg}}{\text{gal}}\right) \left( \frac{4 \text{ gal}}{\text{min}}\right)-
\left(\frac{y \text{ µg}}{200+2t \text{ gal}}\right)\left( \frac{2 \text{ gal}}{\text{min}}\right)$$

@@

$$\frac{1}{t+100}\int 75.708(t+100)dt = \frac{75.708}{t+100} \left( \frac{t^2}{2}+100t+C_1\right)= \frac{75.708}{t+100}\left( \frac{t^2}{2}+100t+C_1\right) $$

@@

$$y(0)=\frac{C_2}{100}$$
```{r}
200*2*c #200 gallons * 2 ppb * ~3.785 (µg/gal)/(1 ppb) = initial amount of dioxin, in µg = y_0
200*2*c*100 #C_2
```

@@

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

@@

(a) $\frac{dy}{dt}=t^2-y^2$. Rewriting as $\frac{dy}{dt}+y^2=t^2$, this is a first order nonlinear ODE.

(b) $17y''=y'+t^2-3y$. Rewriting as  $17y''-y'+3y=t^2$, this is a second order linear ODE with constant coefficients.

(c) $(\sin t)y= \frac{d^3y}{dt^3}- \frac{d^2y}{dt^2}+\cos t$. Rewriting as $\frac{d^3y}{dt^3}- \frac{d^2y}{dt^2}-(\sin t)y=-\cos t$, this is a third order linear ODE with non-constant coefficients.

(d) $yy'=14e^t$ is a first order nonlinear ODE.

@@

Since the patient has had no previous drug treatments, we can assume that there is no drug in the bloodstream at time $t=0$, i.e., we have an initial condition of $y(0)=0$.

@@

$$y(t)= 1250+C_2e^{-0.4t}$$
Plugging in our initial condition,
$0=1250+C_2e^0 \quad\longrightarrow\quad C_2=-1250$

$$y(t)= 1250-1250e^{-0.4t}$$

Check by differentiating:
$$y'(t)=-0.4\left(-1250e^{-0.4t}\right)=500e^{-0.4t}$$
Now we rewrite our initial DE:
$$500-0.4y=$$
$$500-0.4 \left(1250-1250e^{-0.4t}\right)=$$
$$500-500+500e^{-0.4t}=$$
$$500e^{-0.4t}=y'(t)$$
So we have confirmed that out solution satisfies the original differential equation.

@@

## Question 3
$$ \frac{dy}{dx}= \frac{-y}{x}+2, \quad y(1)=3 $$

@@

Check by differentiating:
$$p'(t)=(t+1)(4t)+(2t^2+3)=4t^2+4t+2t^2+3 $$
And since 
$$ 2t^2+3 =\frac{(t+1)(2t^2+3)}{t+1} = \frac{p(t)}{t+1}$$
we get $p'(t)=\frac{1}{t+1}p+4t^2+4t$, which exactly matches our original ODE.

@@

```{r, echo=FALSE}
l_per_g <- 0.2641720524 # 1 L = 0.2641720524 gal
c <- 1/l_per_g
ppb5 <- 5*c

print(paste("1 part per billion = 1 µg/L = 1 µg per", l_per_g,"gal = ", c, "µg/gal"))
print(paste("5 ppb =", ppb5, "µg/gal"))
```

@@

$$\left(\frac{18.927 \text{ µg}}{\text{gal}}\right) \left( \frac{4 \text{ gal}}{\text{min}}\right)-
\left(\frac{y(t) \text{ µg}}{200+2t \text{ gal}}\right)\left( \frac{2 \text{ gal}}{\text{min}}\right)$$

@@

$$\frac{1}{t+100}\int 75.708(t+100)dt = \frac{75.708}{t+100} \left( \frac{t^2}{2}+100t+C_1\right)$$

@@

$$y(0)=\frac{C_2}{100}$$

The starting amount of dioxin, in µg $(y_0)$, is:
```{r}
200*2*c #200 gallons * 2 ppb * 3.785 (µg/gal)/(1 ppb)

```
Which means that $C_2$ is 151416.5 and the solution to the IVP is

@@

Check our solution by differentiating:
$$y'(t)=\frac{(t+100)[(37.9)(2t)+7570.8]-(37.9t^2+7570.8t+151416.5)}{(t+100)^2}=$$
$$\frac{75.708t^2+7570.8t+7570.8t+757082-(37.9t^2+7570.8t+151416.5)}{(t+100)^2}= $$

$$\frac{75.708t^2+15141.6t+757082-(37.9t^2+7570.8t+151416.5)}{(t+100)^2}$$

Our original ODE was

$$\frac{dy}{dt}= 75.708-\frac{y}{100+t} = $$

$$75.708- \frac{37.9t^2+7570.8t+151416.5}{(t+100)^2}=$$

$$75.708 \frac{(t+100)^2}{(t+100)^2}- \frac{37.9t^2+7570.8t+151416.5}{(t+100)^2}=$$

$$\frac{75.708(t+100)^2 - (37.9t^2+7570.8t+151416.5)}{(t+100)^2}=$$

$$\frac{75.708t^2+15141.6t+757082-(37.9t^2+7570.8t+151416.5)}{(t+100)^2}$$
Which is exactly what we got by differentiating our solution.

However, the question does not ask for the solution to the IVP, it asks how much dioxin is in the tank when the tank is full. The tank will be full at the time $t$ such that $200+2t=400$, which means the tank is full at $t=100$ minutes. Now we can plug $t=100$ into our solution:

$$y(100)=\frac{379000+757080+151416.5}{200}=6437.5 \text{ µg} =6.44 \text{ mg}$$

```{r}
# Using unrounded values gives us the same answer, 6.44 mg
((c*10*100*100)+(c*2*1000*100)+(c*4*10000))/200/1000

# Divide by 400 gal and convert units to ppb
((c*10*100*100)+(c*2*1000*100)+(c*4*10000))/200/(400*c)

```

So, there are approximately 6.44 mg of dioxin in the tank when it is full (containing 400 gallons), corresponding to around 4.25 parts per billion. Given that the initial concentration was 2 parts per billion and the water entering the tank is 5 parts per billion, this seems like a reasonable value.
